---
title: "Working Time graph"
format: html
author: Owen James-Gray
---

## DOES WORKING A JOB IMPACT STUDENT PERFORMANCE?

### PURPOSE

Looking through the questions asked on the PISA 2022 dataset I saw that the number of times a student had worked a job during the week was included. I was interested to see if this had an effect on their scores in the tested domains of reading, maths and science. I had a couple of hypothesis to check that work in opposite directions.

### *Hypothesis 1 - A Positive Correlation*

Working in free time during non-school hours displays a strong work ethic and hence that same work ethic could be applied to school work too.

### *Hypothesis 2 - A Negative Correlation*

Working in their free time has an opportunity cost, every hour spent working is an hour that could have been spent studying, resulting in a negative impact on PISA score.

### *The Statistical Methodology*

Using R I took the PISA 2022 dataset, averaged the three scores (reading, maths and science) of each pupil to get a total score for each pupil (SCORE_AVG).

I realised that certain countries with different levels of economic development and different cultures surrounding children working may also have different PISA scores so I came up with a modified pupil score relative to the scores of their own country, this score represents the number of standard deviations a student is away from the average score of their own country (STU_REL_SCORE) so for example a value of -0.55 would mean that student had an average score in the three domains that was 0.55 standard deviations below the country average score. This was then compared against the number of times a student had worked during the week.

Here is the code:

```{r}
library(arrow)
library(tidyverse)

# PISA_2022 <- read_parquet(r"[C:\Users\owenp\OneDrive\Documents\R\PISA_student_2022_subset.parquet]")

PISA_2022 <- read_parquet("/Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/2022/PISA_student_2022_subset.parquet")

avg_score <- PISA_2022 %>% 
  select(CNT, PV1MATH, PV1READ, PV1SCIE, WORKPAY) %>% 
  mutate(SCORE_AVG = ((PV1MATH + PV1READ + PV1SCIE)/3)) %>% 
  group_by(CNT) %>% 
  mutate(CNT_AVG_SCORE = mean(SCORE_AVG)) %>%
  mutate(CNT_SD_SCORE = sd(SCORE_AVG)) %>% 
  ungroup() %>% 
  mutate(STU_REL_SCORE = ((SCORE_AVG-CNT_AVG_SCORE)/CNT_SD_SCORE)) %>% 
  select(CNT, SCORE_AVG, CNT_AVG_SCORE, CNT_SD_SCORE, STU_REL_SCORE, WORKPAY) %>%
  filter(!is.na(WORKPAY))
```

When plotted this results in the following graph:

```{r}
# Define a custom function to map original levels to shorter labels
map_workpay_labels <- function(original_label) {
  # Define mapping for each level
  level_mapping <- c(
    "No work for pay" = "zero",
    "1 time of working for pay per week" = "one",
    "2 times of working for pay per week" = "two",
    "3 times of working for pay per week" = "three",
    "4 times of working for pay per week" = "four",
    "5 times of working for pay per week" = "five",
    "6 times of working for pay per week" = "six",
    "7 times of working for pay per week" = "seven",
    "8 times of working for pay per week" = "eight",
    "9 times of working for pay per week" = "nine",
    "10 or more times of working for pay per week" = "more than ten"
  )
  
  # Use the level_mapping to map the original_label to the shorter label
  return(level_mapping[original_label])
}

# Apply the custom function to the "WORKPAY" column
avg_score$WORKPAY <- sapply(avg_score$WORKPAY, map_workpay_labels)

# Convert "WORKPAY" to a factor with desired order of levels
avg_score$WORKPAY <- factor(avg_score$WORKPAY, levels = c("zero", "one", "two", "three", "four", "five", "six", "seven", "eight", "nine", "more than ten"))

ggplot(data = avg_score,
       aes(x=WORKPAY, y=STU_REL_SCORE)) +
  geom_violin(fill="lightblue")+
  ggtitle("How working affects student scores across the world")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  xlab("Number of times worked per week") +
  ylab("Student average score normalised relative to own country") +
  stat_summary(fun = "mean", geom = "point", color = "black",
               position = position_dodge(width=0.9))

```

In which there does seem to be a negative correlation between number of times a student works and their relative score.

This is made clearer if we remove the violin distribution plots and just look at the averages. The times worked per week as contained within the PISA dataset is a categoric variable so in order to do a linear plot it needs to be made into a numeric variable first so the little bit of code below does that before then plotting a line graph.

```{r}
avg_score <- PISA_2022 %>% 
  select(CNT, PV1MATH, PV1READ, PV1SCIE, WORKPAY) %>% 
  mutate(SCORE_AVG = ((PV1MATH + PV1READ + PV1SCIE)/3)) %>% 
  group_by(CNT) %>% 
  mutate(CNT_AVG_SCORE = mean(SCORE_AVG)) %>%
  mutate(CNT_SD_SCORE = sd(SCORE_AVG)) %>% 
  ungroup() %>% 
  mutate(STU_REL_SCORE = ((SCORE_AVG-CNT_AVG_SCORE)/CNT_SD_SCORE)) %>% 
  select(CNT, SCORE_AVG, CNT_AVG_SCORE, CNT_SD_SCORE, STU_REL_SCORE, WORKPAY) %>%
  mutate(WORKPAY = 
           case_when(WORKPAY == "No work for pay" ~ 0,
                     WORKPAY == "1 time of working for pay per week" ~ 1,
                     WORKPAY == "2 times of working for pay per week" ~ 2,
                     WORKPAY == "3 times of working for pay per week" ~ 3,
                     WORKPAY == "4 times of working for pay per week" ~ 4,
                     WORKPAY == "5 times of working for pay per week" ~ 5,
                     WORKPAY == "6 times of working for pay per week" ~ 6,
                     WORKPAY == "7 times of working for pay per week" ~ 7,
                     WORKPAY == "8 times of working for pay per week" ~ 8,
                     WORKPAY == "9 times of working for pay per week" ~ 9,
                     WORKPAY == "10 or more times of working for pay per week" ~ NA,
                     .default = NA)) %>%
  filter(!is.na(WORKPAY))

ggplot(data = avg_score,
       aes(x=WORKPAY, y=STU_REL_SCORE)) +
  geom_smooth(method = 'lm')+
  ggtitle("How working affects student scores across the world")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  xlab("Number of times worked per week") +
  ylab("Student average score normalised relative to own country") +
  stat_summary(fun = "mean", geom = "point", color = "black",
               position = position_dodge(width=0.9))
```

Which does show more clearly a negative correlation between times worked per week when times is less than 5. The correlation gets a little more noisy after that. I speculate this is because working more than 5 times per week would mean working more than every weekday, which might mean students answering the questionnaire with larger numbers may be defining a "working" period differently than those answering with smaller values or perhaps have misread the question making the data noisier for values over 5.

One thing in discussion with Peter and Richard I realised was that this model while it takes account of differences in economic status by country, it does not take into account differences in socioeconomic status between students. It is completely plausible that even within a country poorer students might be having to work more than richer ones, and it fairly established that students with lower socioeconomic status get lower scores on PISA tests (OECD, 2018). This might mean that all we are seeing in this graph is that effect, rather than any effect from working in a job. To remedy this we can adapt our model to control for socioeconomic status.

### CONTROLING FOR SOCIOECONOMIC STATUS (ESCS)

First of all let's see if there is a relationship between a student working in a job and their ESCS.

```{r}
avg_score <- PISA_2022 %>% 
  select(WORKPAY, ESCS) %>% 
  mutate(WORKPAY = 
           case_when(WORKPAY == "No work for pay" ~ 0,
                     WORKPAY == "1 time of working for pay per week" ~ 1,
                     WORKPAY == "2 times of working for pay per week" ~ 2,
                     WORKPAY == "3 times of working for pay per week" ~ 3,
                     WORKPAY == "4 times of working for pay per week" ~ 4,
                     WORKPAY == "5 times of working for pay per week" ~ 5,
                     WORKPAY == "6 times of working for pay per week" ~ 6,
                     WORKPAY == "7 times of working for pay per week" ~ 7,
                     WORKPAY == "8 times of working for pay per week" ~ 8,
                     WORKPAY == "9 times of working for pay per week" ~ 9,
                     WORKPAY == "10 or more times of working for pay per week" ~ NA,
                     .default = NA)) %>%
  filter(!is.na(WORKPAY))

ggplot(data = avg_score,
       aes(x=WORKPAY, y=ESCS)) +
  geom_smooth(method = 'lm')+
  ggtitle("Do low socioeconomic status students work more times a week?")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  xlab("Number of times worked per week") +
  ylab("Student socioeconomic status") +
  stat_summary(fun = "mean", geom = "point", color = "black",
               position = position_dodge(width=0.9))

```

### REFERENCES

OECD. (2018). Equity in Education: Breaking Down Barriers to Social Mobility. OECD. https://doi.org/10.1787/9789264073234-en
